msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-05 paper.tex
Moved multiple merger derivations to the appendix.

\subsection*{Multiple merger coalescents}
%%Kingman's coalescent assumes that all mergers are binary. This
%%assumption can be violated in some situations. Lambda coalescents
%%~\citep{donnelly1999particle,schweinsberg2000coalescents}
%%HybridLambda~\citep{zhu2015hybrid}
%%We support X Y Z.

\label{mmc}%

@@

In \msprime we have incorporated two examples of multiple-merger
coalescents.  One is a diploid extension \citep{BBE13} of the haploid
model of sweepstakes reproduction considered by \cite{EW06}, which is
a haploid Moran model adapted to sweepstakes reproduction.  Let $N$
denote population size, and take $\psi \in (0,1]$ to be fixed.  In
every generation, with probability $1-\varepsilon_N$ a single
individual (picked uniformly at random) perishes, and one of the
surviving individuals (sampled uniformly at random) produces one
offspring; with probability $\varepsilon_N$ a total of
$\lfloor \psi N \rfloor$ individuals perish, and of the remaining
individuals a single individual produces $\lfloor \psi N \rfloor -1 $
offspring.  Taking $\varepsilon_N = 1/N^\gamma$ for some $\gamma > 0$,
\cite{EW06} obtain specific examples of $\Lambda$-coalescents, where
the $\Lambda$ measure in Eq \eqref{lambdabk} is a point mass at
$\psi$.  The simplicity of this model does allow one to obtain some
explicit mathematical results (see e.g.\
\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model
considered by \cite{EW06} has also been applied in algorithms for
simulating gene genealogies within phylogenies \citep{zhu2015hybrid}. The
 specific model incorporated into \msprime is the diploid version
\citep{BBE13} of the model studied by \cite{EW06}, which would be
necessary in order to incorporate recombination.  In the model
considered by \cite{BBE13}, a single pair of diploid individuals
contribute offspring in each generation, selfing is excluded, and each
offspring is assigned one chromosome from each parent. There are,
therefore, four parent chromosomes involved in each reproduction
event, which can lead to up to four simultaneous mergers.  Let
\begin{esplit}{Cconst}
C_{b; k_1, \ldots, k_r;s } &=  \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell} (4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots + k_r + \ell},  \\
\end{esplit}
where  $C_{b; k_1, \ldots, k_r;s }$ corresponds to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq. \eqref{xi}) of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$.
The total merger rate (see Eq.\ \eqref{lambdabkall}) is then
\be\label{xidirlambdabk}
      \lambda(b;k_1, \ldots, k_r) =    \mathcal{N}(b; k_1, \ldots, k_r ) \left( \frac{c\psi^2/4}{1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s } +     \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right),
\ee
and the total coalescence rate (see Eq.\ \eqref{xilambdab}) becomes
    \begin{esplit}{xilambdab}
\lambda_b & =  \frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0 }^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 - \psi)^{b-\ell } \right) +    \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.  \\
\end{esplit}
The interpretation of Eq.\ \eqref{xilambdab} is that with probability  $1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 + c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair produce  $\lfloor \psi N \rfloor$ offspring.

The other multiple-merger coalescent model incorporated in \msprime is
derived from an adaptation of the haploid population model considered
by \cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the
haploid version, in each generation individuals independently produce
random numbers of juveniles, where the random  number of juveniles $(X)$ produced by
a given individual  has the stable law \be\label{jX} \lim_{k\to
\infty} C k^\alpha \prb{X \ge k} = 1 \ee with index $\alpha > 0$, and
$C > 0$ is a normalising constant.   One can interpret  Eq.\ \eqref{jX} as  stating the form of the  probability distribution  for number of juveniles at least on the order of the population size.     If the random  total number of juveniles $(S_N)$ produced in this way is at least the population size $(N)$, then one samples  $N$ juveniles uniformly at random without replacement to form the next generation of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the same  set of individuals.   However,   assuming  $\EE{X} > 1$, one can show that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.         If $\alpha \ge 2$ the ancestral
process converges to the Kingman-coalescent; if $1 \le \alpha < 2$ the
ancestral process converges to a specific case of a
$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.\
\eqref{lambdabk} is associated with the Beta$(2-\alpha, \alpha)$
probability distribution, i.e.\
\be\label{Fbeta}
    \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)} x^{1 - \alpha}(1-x)^{\alpha - 1}  dx,
\ee
where $B(2-\alpha,\alpha)$ is the beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$ \citep{schweinsberg03}.    This model has been adapted to diploid populations  by                      \cite{BLS15}, where  the resulting coalescent process  is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$, where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.  The merger rate (see Eq.\ \eqref{xi}) is then ($1 \le r \le 4$)
\be\label{xibeta}
   \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b - k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)}
\ee
\citep{Blath2016,BLS15}. The interpretation of Eq.\ \eqref{xibeta} is  that the random   number of diploid  juveniles each  diploid pair of parents  produces  is  governed by  the law in Eq.\ \eqref{jX},   and   each diploid juvenile  is assigned  one chromosome  from each parent (selfing is excluded).

The model in Eq.\ \eqref{jX} assumes that individuals can produce
arbitrarily large numbers of juveniles. Considering diploid juveniles,
this assumption is probably rather strong, since diploid juveniles are
at least fertilised eggs, and so it is reasonable to suppose that the
number of juveniles surviving to the  particular life stage we are modeling  cannot be
arbitrarily large.  With this in mind, we also consider an adaptation
of the Schweinsberg model, where the random number of juveniles $(X)$
produced by a   given parent pair  is distributed according to
\be\label{jtr}
  \prb{X=k} =   \one{1 \le k \le \phi(N)} \frac{\phi(N+1)^\alpha }{ \phi(N+1)^\alpha - 1 }  \left( \frac{1}{k^\alpha} - \frac{1}{(k+1)^\alpha}  \right) ,
\ee
where $\phi(N)$ is a deterministic strict upper bound on the number of juveniles produced by  any given parent pair (see also \citep{Eldon2018}).   One can follow the calculations in  \citep{schweinsberg03} or \citep{BLS15}  to show  that  if $1 < \alpha < 2$   then  $(k = k_1 + \cdots + k_r)$ then the merger rate (see Eq.\ \eqref{xibeta}) is
\be
   \lambda_{b;k_1, \ldots, k_r} =  \sum_{\ell = 0}^{ (b - k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(M; k+\ell - \alpha, b-k-\ell + \alpha ) }{B(M;2-\alpha,\alpha)}
\ee
with $B(z;a,b) := \int_0^z t^{a-1}(1-t)^{b-1}dt$ for  $a,b>0$ and $0< z\le 1$, and
\be
M :=  \frac{K}{K+m} \one{\phi(N) = KN} + \one{\phi(N)/N \to \infty }
\ee
where $K > 0$ is a constant and  $m := \lim_{N\to \infty} \EE{X} = 1 + 2^{1-\alpha}/(\alpha - 1)$ \citep{CDEE2020,AEKKZ2020}.    In other words,  the measure $(\Lambda)$ driving the multiple mergers is of the same form as in Eq.\ \eqref{Fbeta}  with $0 < x \le M$ in the case $1 < \alpha < 2$ and $\phi(N) \ge KN$.   If $\alpha \ge 2$ or $\phi(N)/N \to 0$ then  the  ancestral process converges (in the sense of convergence of  finite-dimensional distributions) to  the Kingman-coalescent \citep{CDEE2020,AEKKZ2020}.

@@

Mention~\citep{becheler2020occupancy} somewhere.

@@

>>>>

\subsection*{Multiple merger coalescents}

@@

@@

@@

\appendix
\label{app-multiple-mergers}
\section{Multiple merger coalescent model}

In \msprime\ we have incorporated two examples of multiple-merger coalescents.
One is a diploid extension \citep{BBE13} of the haploid model of sweepstakes
reproduction considered by \cite{EW06}, which is a haploid Moran model adapted
to sweepstakes reproduction.  Let $N$ denote population size, and take $\psi
\in (0,1]$ to be fixed.  In every generation, with probability
$1-\varepsilon_N$ a single individual (picked uniformly at random) perishes,
and one of the surviving individuals (sampled uniformly at random) produces one
offspring; with probability $\varepsilon_N$ a total of $\lfloor \psi N \rfloor$
individuals perish, and of the remaining individuals a single individual
produces $\lfloor \psi N \rfloor -1 $ offspring.  Taking $\varepsilon_N =
1/N^\gamma$ for some $\gamma > 0$, \cite{EW06} obtain specific examples of
$\Lambda$-coalescents, where the $\Lambda$ measure in Eq.~\eqref{lambdabk} is a
point mass at $\psi$.  The simplicity of this model does allow one to obtain
some explicit mathematical results (see e.g.\
\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model considered by
\cite{EW06} has also been applied in algorithms for simulating gene genealogies
within phylogenies \citep{zhu2015hybrid}. The specific model incorporated into
\msprime\ is the diploid version \citep{BBE13} of the model studied by
\cite{EW06}, which would be necessary in order to incorporate recombination.
In the model considered by \cite{BBE13}, a single pair of diploid individuals
contribute offspring in each generation, selfing is excluded, and each
offspring is assigned one chromosome from each parent. There are, therefore,
four parent chromosomes involved in each reproduction event, which can lead to
up to four simultaneous mergers.  Let \begin{esplit}{Cconst} C_{b; k_1, \ldots,
k_r;s } &= \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell}
(4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots +
k_r + \ell},  \\ \end{esplit} where  $C_{b; k_1, \ldots, k_r;s }$ corresponds
to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq~\eqref{xi})
of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$. The
total merger rate (see Eq~\eqref{lambdabkall}) is then
\be\label{xidirlambdabk} \lambda(b;k_1, \ldots, k_r) = \mathcal{N}(b; k_1,
\ldots, k_r ) \left( \frac{c\psi^2/4}{ 1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s
} + \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right), \ee and the total
coalescence rate becomes \begin{esplit}{xilambdab} \lambda_b & =
\frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0
}^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 -
\psi)^{b-\ell } \right) + \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.  \\
\end{esplit} The interpretation of \eqref{xilambdab} is that with probability
$1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent
pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 +
c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair
produce  $\lfloor \psi N \rfloor$ offspring.

The other multiple-merger coalescent model incorporated in \msprime\ is derived
from an adaptation of the haploid population model considered by
\cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the haploid
version, in each generation individuals independently produce random numbers of
juveniles, where the random  number of juveniles $(X)$ produced by a given
individual  has the stable law \be\label{jX} \lim_{k\to \infty} C k^\alpha
\prb{X \ge k} = 1 \ee with index $\alpha > 0$, and $C > 0$ is a normalising
constant.   One can interpret  Eq.~\eqref{jX} as  stating the form of the
probability distribution  for number of juveniles at least on the order of the
population size.     If the random  total number of juveniles $(S_N)$ produced
in this way is at least the population size $(N)$, then one samples  $N$
juveniles uniformly at random without replacement to form the next generation
of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the
same  set of individuals.   However,   assuming  $\EE{X} > 1$, one can show
that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.
If $\alpha \ge 2$ the ancestral process converges to the Kingman-coalescent; if
$1 \le \alpha < 2$ the ancestral process converges to a specific case of a
$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.~\eqref{lambdabk} is
associated with the Beta$(2-\alpha, \alpha)$ probability distribution, i.e.\
\be\label{Fbeta} \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)}
x^{1 - \alpha}(1-x)^{\alpha - 1}  dx, \ee where $B(2-\alpha,\alpha)$ is the
beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$
\citep{schweinsberg03}.
This model has been adapted to diploid populations
by \cite{BLS15}, where  the resulting coalescent process
is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$,
where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.
The merger rate (see Eq.~\eqref{xi}) is then ($1 \le r \le 4$)
\be\label{xibeta} \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b -
k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}}
\frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)} \ee
\citep{Blath2016,BLS15}. The interpretation of Eq.~\eqref{xibeta} is  that the
random   number of diploid  juveniles each  diploid pair of parents  produces
is  governed by  the law in Eq.~\eqref{jX},   and   each diploid juvenile  is
assigned  one chromosome  from each parent (selfing is excluded).

The model in Eq~\eqref{jX} assumes that individuals can produce arbitrarily
large numbers of juveniles. Considering diploid juveniles, this assumption is
probably rather strong, since diploid juveniles are at least fertilised eggs,
and so it is reasonable to suppose that the number of juveniles surviving to
the  particular life stage we are modeling  cannot be arbitrarily large.  With
this in mind, we also consider an adaptation of the Schweinsberg model, where
the random number of juveniles $(X)$ produced by a   given parent pair  is
distributed according to \be\label{jtr} \prb{X=k} =   \one{1 \le k \le \phi(N)}
\frac{\phi(N+1)^\alpha }{ \phi(N+1)^\alpha - 1 }  \left( \frac{1}{k^\alpha} -
\frac{1}{(k+1)^\alpha}  \right) , \ee where $\phi(N)$ is a deterministic strict
upper bound on the number of juveniles produced by  any given parent pair (see
also \citep{Eldon2018}).   One can follow the calculations in
\citep{schweinsberg03} or \citep{BLS15}  to show  that  if $1 < \alpha < 2$
then  $(k = k_1 + \cdots + k_r)$ then the merger rate (see Eq.~\eqref{xibeta})
is
 \be \lambda_{b;k_1, \ldots, k_r} =  \sum_{\ell = 0}^{ (b - k)\wedge (4-r) }
\binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(M; k+\ell - \alpha,
b-k-\ell + \alpha ) }{B(M;2-\alpha,\alpha)}
\ee
with $B(z;a,b) := \int_0^z
t^{a-1}(1-t)^{b-1}dt$ for  $a,b>0$ and $0< z\le 1$, and \be M :=  \frac{K}{K+m}
\one{\phi(N) = KN} + \one{\phi(N)/N \to \infty } \ee where $K > 0$ is a
constant and  $m := \lim_{N\to \infty} \EE{X} = 1 + 2^{1-\alpha}/(\alpha - 1)$
\citep{CDEE2020,AEKKZ2020}.    In other words,  the measure $(\Lambda)$ driving
the multiple mergers is of the same form as in Eq.~\eqref{Fbeta}  with $0 < x
\le M$ in the case $1 < \alpha < 2$ and $\phi(N) \ge KN$.   If $\alpha \ge 2$
or $\phi(N)/N \to 0$ then  the  ancestral process converges (in the sense of
convergence of  finite-dimensional distributions) to  the Kingman-coalescent
\citep{CDEE2020,AEKKZ2020}.
